---
title: "ANCOM-BC2 Analysis and Visualization with ancombcVizhelper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ANCOM-BC2 Analysis and Visualization with ancombcVizhelper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 6,
  out.width = "100%"
)
```

## Overview

This vignette demonstrates an end-to-end ANCOM-BC2 workflow using the `atlas1006` example dataset from the `microbiome` package.

The data preparation and ANCOM-BC2 model specification are adapted from the ANCOM-BC2 tutorial. The visualization stage uses `ancombcVizhelper` instead of manually reshaping ANCOM-BC2 result tables for each plot.

The workflow includes:

1. Loading and preparing the `atlas1006` dataset.
2. Running ANCOM-BC2 with BMI as the grouping variable.
3. Inspecting the ANCOM-BC2 result tables.
4. Creating global-test heatmaps and bar plots.
5. Showing code patterns for pairwise, Dunnett-type, and trend results.

## Load and Prepare the Example Data

```{r load-data}
utils::data("atlas1006", package = "microbiome")

ps <- atlas1006

# Baseline samples only
ps <- phyloseq::subset_samples(
  ps,
  time == 0
)

meta <- phyloseq::sample_data(ps) |>
  base::as.data.frame()

meta$bmi <- dplyr::recode(
  meta$bmi_group,
  obese = "obese",
  severeobese = "obese",
  morbidobese = "obese"
)

meta$region <- dplyr::recode(
  base::as.character(meta$nationality),
  Scandinavia = "NE",
  UKIE = "NE",
  SouthEurope = "SE",
  CentralEurope = "CE",
  EasternEurope = "EE",
  .missing = "unknown"
)

meta$bmi <- base::factor(
  meta$bmi,
  levels = c("obese", "overweight", "lean")
)

phyloseq::sample_data(ps) <- meta

# Keep requested BMI groups and valid regions
ps <- phyloseq::subset_samples(
  ps,
  bmi %in% c("obese", "overweight", "lean") &
    !region %in% c("EE", "unknown")
)

# Remove any taxa that became zero after sample filtering
ps <- phyloseq::prune_taxa(
  phyloseq::taxa_sums(ps) > 0,
  ps
)
```

## Run ANCOM-BC2

The tutorial uses a small number of bootstrap iterations for the trend test to reduce execution time. This vignette keeps `B = 10` for demonstration only.

For a final analysis, use the ANCOM-BC2 default or a larger value such as `B = 100` or higher.

```{r run-ancombc2}
set.seed(42)


output <- ANCOMBC::ancombc2(
  data = ps,
  assay_name = "counts",
  tax_level = "Family",
  fix_formula = "age + region + bmi",
  rand_formula = NULL,
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0.10,
  lib_cut = 1000,
  s0_perc = 0.05,
  group = "bmi",
  struc_zero = TRUE,
  neg_lb = TRUE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = TRUE,
  trend = TRUE,
  iter_control = base::list(
    tol = 1e-2,
    max_iter = 20,
    verbose = TRUE
  ),
  em_control = base::list(tol = 1e-5, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = base::list(fwer_ctrl_method = "holm", B = 100),
  trend_control = base::list(
    contrast = base::list(
      base::matrix(c(1, 0, -1, 1), nrow = 2, byrow = TRUE),
      base::matrix(c(-1, 0, 1, -1), nrow = 2, byrow = TRUE),
      base::matrix(c(1, 0, 1, -1), nrow = 2, byrow = TRUE)
    ),
    node = base::list(2, 2, 1),
    solver = "ECOS",
    B = 10
  )
)
```

## Inspect the ANCOM-BC2 Result Object

```{r inspect-results}
base::names(output)
```

The primary result components used by `ancombcVizhelper` are:

| Result object | Purpose |
|---|---|
| `output$res` | Primary regression coefficients and covariate effects |
| `output$res_global` | Taxon-level global test across BMI groups |
| `output$res_pair` | Pairwise directional comparisons with mdFDR control |
| `output$res_dunn` | Dunnett-type comparisons against the reference BMI group |
| `output$res_trend` | Ordered trend-test results |
| `output$zero_ind` | Structural-zero indicators |

## Global-Test Heatmap

The global test first identifies taxa with evidence of an overall difference among BMI groups. The displayed log-fold changes are the BMI coefficients from
`output$res`.

```{r global-heatmap}



global_heatmap <- ancombcVizhelper::make_heatmap(
  out = output,
  result = "res_dunn",
  prefix = "bmi",
  title = "Global ANCOM-BC2 BMI-associated families",
  sensitivity = "keep",
  show_all = FALSE
)

global_heatmap$plot

```

Structural-zero taxa are excluded from log-fold change plots because their
coefficients are not estimated by the standard ANCOM-BC2 model.

```{r structural-zeros}
utils::head(global_heatmap$structural_zero)
```

## Global-Test Bar Plot

Bars show the BMI model coefficients for families selected by the global test. Error bars represent plus or minus one standard error. Positive log-fold changes are red and negative log-fold changes are blue. Fully opaque bars are sensitivity-robust, whereas semi-transparent bars did not pass the pseudo-count sensitivity criterion.

```{r global-barplot}
global_barplot <- ancombcVizhelper::make_barplots(
  out = output,
  result = "res_global",
  prefix = "bmi",
  title = "BMI coefficients for globally significant families",
  sensitivity = "keep",
  show_all = FALSE,
  group_order = "mean",
  order = "asc"
)

global_barplot$plot

```

## Pairwise Directional Comparisons

`res_pair` is appropriate when all pairwise directional contrasts are of interest. The `diff_*` results in this object correspond to mdFDR-controlled pairwise inference.

```{r pairwise-heatmap, eval = FALSE}
pairwise_heatmap <- ancombcVizhelper::make_heatmap(
  out = output,
  result = "res_pair",
  prefix = "bmi",
  title = "Pairwise BMI comparisons",
  sensitivity = "keep",
  show_all = FALSE
)
pairwise_heatmap$plot

```

## Dunnett-Type Comparisons

`res_dunn` is useful when each BMI group is compared with the reference group. In this workflow, the reference group is `obese`.

```{r dunnett-heatmap, eval = FALSE}
dunnett_heatmap <- ancombcVizhelper::make_heatmap(
  out = output,
  result = "res_dunn",
  prefix = "bmi",
  title = "BMI comparisons against obese participants",
  sensitivity = "keep",
  show_all = FALSE
)

dunnett_heatmap$plot

dunnett_barplot <- ancombcVizhelper::make_barplots(
  out = output,
  result = "res_dunn",
  prefix = "bmi",
  title = "BMI comparisons against obese participants",
  sensitivity = "keep",
  show_all = FALSE
)

dunnett_barplot$plot

```

## Trend-Test Results

The trend test identifies taxa showing the specified ordered patterns across
BMI groups. The log-fold changes shown in the heatmap are fitted BMI
coefficients, while significance is determined by the taxon-level trend test.

```{r trend-heatmap, eval = FALSE}
trend_heatmap <- ancombcVizhelper::make_heatmap(
  out = output,
  result = "res_trend",
  prefix = "bmi",
  title = "BMI trend-test results",
  sensitivity = "keep",
  show_all = FALSE
)

trend_heatmap$plot
```

## Primary Result for a Continuous Covariate

For a continuous covariate such as age, `result = "res"` is appropriate because there is one regression coefficient rather than multiple BMI contrasts.

```{r age-barplot, eval = FALSE}
age_barplot <- ancombcVizhelper::make_barplots(
  out = output,
  result = "res",
  prefix = "age",
  title = "Family-level associations with age",
  sensitivity = "keep",
  show_all = FALSE,
  group_order = "mean",
  order = "asc"
)

age_barplot$plot
```

## Interpretation Notes

- Use `res_global` to select taxa based on an overall group difference.
- Use `res_pair` for all pairwise directional comparisons with mdFDR control.
- Use `res_dunn` for comparisons against a predefined reference group.
- Use `res_trend` when an ordered group trend was specified in `ancombc2()`.
- Use `res` primarily for continuous covariates or a single coefficient.
- Use `sensitivity = "robust_only"` when only taxa passing pseudo-count
  sensitivity analysis should be displayed.
